# Data

Sleep data was drawn from the [Center for Disease Control and Prevention](http://www.cdc.gov/places/) and population demographic data from the [U.S. Census Bureau](https://data.census.gov/all/tables).

## Description

### Population Demographics: Race and Socioeconomic Data {.unnumbered}
These data are available on the U.S. Census Bureau website. Make sure to select all ZCTAs that fall within New York state!

* [Race and Ethnicity Data](https://data.census.gov/table/ACSDT5Y2022.B03002?q=Race%20and%20Ethnicity&g=040XX00US36$8600000&y=2022)
* [Selected Demographics (Part 1)](https://data.census.gov/table/ACSDP5Y2022.DP03?q=selected%20demographi&g=860XX00US10002,10003,10004,10006&y=2022)
* [Selected Demographics (Part 2)](https://data.census.gov/table/ACSDP5Y2022.DP02?q=selected%20demographi&g=860XX00US10002,10003,10004,10006&y=2022)

### Sleep Data Over Time {.unnumbered}
Sleep data are obtained from the CDC's PLACES Local Data for Better Health datasets directly from the CDC website (Center for Disease Control and Prevention, 2020, 2023, 2024). The PLACES datasets are created with ZCTA-level estimates derived from data collected through the nationally representative annual U.S. Behavioral Risk Factor Surveillance Survey (BRFSS). The BRFSS is an annual, phone-based survey of the US population who are 18 years and older. The CDC works with local state health department to interview respondents. Every other year, the BRFSS includes a question asking "On average, how many hours of sleep do you get in a 24-hour period?". 

Using the BRFSS data, an estimate is created for the percentage of the population (per ZCTA) that reports sleeping < 7 hours on average. Population data used to help model these estimates were drawn from the American Community Surveys (ACS) for each of the respective BRFSS years (2018, 2020, 2022). The ACS is collected and managed by the U.S. Census Bureau. Because sleep data is only collected every other year and PLACES short sleep prevalence estimates are created using the most recent year with available data (even if the year the data were collected does not align with the release year of the PLACES dataset).

### Additional Resources {.unnumbered}

::: {.callout-note}
## Data Sources
* [NYC ZCTA to Borough to Neighborhood Crosswalk File](https://data.cityofnewyork.us/Health/Modified-Zip-Code-Tabulation-Areas-MODZCTA-/pri4-ifjk/data_preview)
* [ZCTA Shapefile](https://nychealth.github.io/covid-maps/modzcta-geo/about.html)
:::

::: {.callout-tip}
## Data Cleaning
All data cleaning files and steps can be found in the [data_cleaning folder](https://github.com/cj-concepcion/sleep-health/tree/main/data_cleaning) *(BONUS: a funny joke and pun awaits inside if you can find it)*
:::

## Missing value analysis

Because each row of the sleep data is a model estimated value, there actually aren't any NAs in this dataset, this is demonstrated by the plot below.

### Missing Value Plot
```{r}
# Load required libraries
library(tidyverse)
library(reshape2)

# Read the data
df <- read.csv("data_cleaning/sleep_health_final_dataset.csv")

# Create missing value matrix
missing_matrix <- is.na(df) |> 
  as.data.frame() |>
  mutate(row_id = row_number()) |>
  melt(id.vars = "row_id", variable.name = "column", value.name = "is_missing")

# Create the heatmap
ggplot(missing_matrix, aes(x = column, y = row_id, fill = is_missing)) +
  geom_tile() +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "salmon")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Missing Values Heatmap",
       x = "Variables",
       y = "Row Number",
       fill = "Is Missing") +
  coord_flip()
```

## Zero Value Analysis
```{r}
# Calculate percentage of zeros for all numeric columns
zero_summary <- sapply(df, function(x) if(is.numeric(x)) mean(x == 0) * 
                         100 else NA) |>
  as.data.frame() |>
  setNames("zero_percentage") |>
  rownames_to_column("variable") |>
  filter(!is.na(zero_percentage), zero_percentage > 0) |>
  arrange(desc(zero_percentage))

# Create a bar plot
ggplot(zero_summary, aes(x = reorder(variable, zero_percentage), 
                         y = zero_percentage)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = sprintf("%.1f%%", zero_percentage)), 
            hjust = -0.1, 
            size = 3.5) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Percentage of Zero Values by Column",
       x = "Variables",
       y = "Percentage of Zeros (%)") +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 10),
    panel.grid.major.y = element_blank()
  ) +
  scale_y_continuous(limits = c(0, max(zero_summary$zero_percentage) * 1.2))
```

- Our highest instances of zero values are for percentage of native populations (Native Hawaiian or Other Pacific Islander and American Indian or Alaska Native). This likely reflects true 0 values and indicates the under representation of these groups within New York City.

```{r}
# Create a better visualization of the zero patterns
zero_locations <- df |>
  select(zcta5, edu_gr9, inc_15k, inc_25k) |>
  filter(edu_gr9 == 0 | inc_15k == 0 | inc_25k == 0)

# Convert data to long format for better visualization
zero_long <- zero_locations |>
  pivot_longer(cols = c(edu_gr9, inc_15k, inc_25k),
               names_to = "variable",
               values_to = "value") |>
  mutate(is_zero = value == 0)

# Create a more readable visualization
ggplot(zero_long, aes(x = factor(zcta5), y = variable, fill = is_zero)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("steelblue", "coral")) +
  theme_minimal() +
  labs(title = "Zero Values Pattern Across ZCTAs",
       x = "ZCTA",
       y = "Variable",
       fill = "Is Zero") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()) +
  scale_x_discrete(name = "ZCTA")

# Create a summary table of the patterns
zero_summary <- zero_long |>
  group_by(zcta5) |>
  summarise(
    num_zeros = sum(is_zero),
    variables_with_zeros = paste(variable[is_zero], collapse = ", ")
  )

zero_summary
```

- The systematic pattern of zero values in certain ZCTAs (especially Manhattan's 10004, 10006 and other Financial District areas) likely reflects extreme income stratification - these wealthy neighborhoods simply have no households in the lower income brackets (under 15k and 25k), rather than indicating missing data.
